### 2023-08-30 misc/expo/kelleymeka.tex
More kelley meka

We will use $G$ primarily to refer to a finite abelian group.
For functions $f,g:G\to \CC$ we have the inner product
$$\langle f,g\rangle = \ex_{x\in G} f(x)\bar{g(x)}$$
and the $L_p$ norm
$$ \norm{f}_p = \Bigl( \ex_{x\in G} \bigl| f(x) \bigr|^p \Bigr)^{1/p}.$$
In $L_p$ spaces we have the useful H\"older inequality
$$ \bigl| \langle f,g\rangle \bigr| \le \norm{f}_p  \cdot \norm{g}_q,$$
for $p,q\in [1,\infty)$ with $1/p + 1/q = 1$.
Assuming now that $f$ and $g$ are $\RR$-valued, we also have the convolution
$$ (f*g)(x) = \ex_{y\in G} f(y) g(x-y)$$
and the difference convolution
$$ (f\circ g)(x) = \ex_{y\in G} f(y) g(x+y).$$
It is easy to check that for all $x\in G$, $(f*g)(x) = (g*f)(x)$, but with the difference convolution
we have $(f\circ g)(x) = (g\circ f)(-x)$.
that are related by the following adjoint property.

@@

For a group $G$ the dual group $\hat G$ is the set of all homomorphisms from $G\to \CC^\times$.
The Fourier transform of $f:G\to \RR$ is the function $\hat f:\hat G\to \CC$ given by
$$ \hat f(\chi) = \ex_{x\in G} f(x)\chi(-x).$$
The following proposition describes how the convolution and difference convolution behave under the Fourier
transform.

@@

\advsect H\"older lifting and unbalancing

With preliminaries out of the way, we begin the proof of Kelley and Meka~\ref{kelleymeka2023}, as described
and reworked by Bloom and Sisask~\ref{bloomsisask2023}.
In this section we perform the first two steps of the proof,
which are identical in both the integer and finite-field settings.

@@

There is a theorem that we need which, for now, we shall just state without proof. (This is Theorem 3.2
of~\ref{schoensisask2016}.)
[TODO: Add whole section proving this.]

@@

\parenproclaim Lemma~{\advthm} (Density increment). Let $\eps\in (0,1)$ and let $A$ and $C$
be subsets of $G =\FF_q^n$ with relative densities $\alpha$ and $\gamma$, respectively. Then either
\medskip
\item{i)} $\bigl|\langle \mu_A*\mu_A,\mu_C\rangle-1\bigr| \le \eps$; or
\smallskip
\item{ii)} there is a subspace $V$ of codimension
$$O\Bigl(\eps^{-2}\bigl(\log(1/\gamma)+\eps^{-1}\log(\eps^{-1})\bigr)^4\log(1/\alpha)^4\Bigr)$$
such that $\max_{x\in G}\;(\one_A*\mu_V)(x) \ge \bigl(1+\eps/32\bigr)\alpha$.
\medskip

@@

>>>>

We will use $G$ primarily to refer to a finite abelian group.
For functions $f,g:G\to \CC$ we have the inner product
$$\langle f,g\rangle = \ex_{x\in G} f(x)\bar{g(x)}$$
and the $L_p$ norm
$$ \norm{f}_p = \Bigl( \ex_{x\in G} \bigl| f(x) \bigr|^p \Bigr)^{1/p}.$$
In $L_p$ spaces we have the useful H\"older inequality
$$ \bigl| \langle f,g\rangle \bigr| \le \norm{f}_p  \cdot \norm{g}_q,$$
for $p,q\in [1,\infty)$ with $1/p + 1/q = 1$.
Assuming now that $f$ and $g$ are $\RR$-valued, we also have the convolution
$$ (f*g)(x) = \ex_{y\in G} f(y) g(x-y)$$
and the difference convolution
$$ (f\circ g)(x) = \ex_{y\in G} f(y) g(x+y).$$
It is easy to check that for all $x\in G$, $(f*g)(x) = (g*f)(x)$, but with the difference convolution
we have $(f\circ g)(x) = (g\circ f)(-x)$.
We also have the following adjoint property.

@@

For a group $G$ the dual group $\hat G$ is the set of all homomorphisms from $G$ to $\CC^\times$.
The Fourier transform of $f:G\to \RR$ is the function $\hat f:\hat G\to \CC$ given by
$$ \hat f(\chi) = \ex_{x\in G} f(x)\chi(-x).$$
The following proposition describes how the convolution and difference convolution behave under the Fourier
transform.

@@

\advsect H\"older lifting and unbalancing for finite groups

With preliminaries out of the way, we begin the proof of Kelley and Meka~\ref{kelleymeka2023}, as described
and reworked by Bloom and Sisask~\ref{bloomsisask2023}.
In this section we perform the first two steps of the proof, in the general setting of finite groups.

@@

There is a theorem that we need which, for now, we shall just state without proof. (This is Theorem 3.2
of~\ref{schoensisask2016}.)

@@

\edef\densityincrement{\the\thmcount}
\parenproclaim Lemma~{\advthm} (Density increment). Let $\eps\in (0,1)$ and let $A$ and $C$
be subsets of $G =\FF_q^n$ with relative densities $\alpha$ and $\gamma$, respectively. Then either
\medskip
\item{i)} $\bigl|\langle \mu_A*\mu_A,\mu_C\rangle-1\bigr| \le \eps$; or
\smallskip
\item{ii)} there is a subspace $V$ of codimension
$$O\Bigl(\eps^{-2}\bigl(\log(1/\gamma)+\eps^{-1}\log(\eps^{-1})\bigr)^4\log(1/\alpha)^4\Bigr)$$
such that $\max_{x\in G}\;(\one_A*\mu_V)(x) \ge \bigl(1+\eps/32\bigr)\alpha$.
\medskip

@@

\parenproclaim Theorem~{\advthm} (Finite field). Let $q$ be a power of an odd prime and let $A$ be
a subset of $G = \FF_q^n$ of cardinality $\alpha q^n$. The number of (possibly trivial)
three-term arithmetic progressions contained in $A$ is at least
$${\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)}.$$
Hence if $A\subseteq \FF_q^n$ contains no nontrivial three-term arithmetic progressions, then
$\alpha \le e^{-\Omega(n^{1/9})} = O\bigl(q^{-\Omega(n^{1/9})}\bigr)$.

\proof Let $C = 2\cdot A = \{2a : a\in A\}$, so that $\gamma = |C|/q^n = \alpha$. By Lemma~{\densityincrement}
applied to $A$ and $C$ with parameter $\eps = 1/2$, we find that either
$\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$ or there is a subspace $V$ of codimension
$O\bigl(\log(1/alpha)^8\bigr)$ such that $\max_{x\in G}\;(\one_A*\mu_V)(x)\ge (1+\eps/64)\alpha$.

In the second case, there exists some $x\in G$ such that
$$\ex_{y\in G} \one_A(y)\mu_V(x-y) \ge (1+\eps/64)\alpha.$$
But $x-y\in V$ if and only if $y-x\in V$ if and only if $y\in V+x$, so we find that
$$|A\cap V+x| \ge (1+\eps/64)\alpha|V|.$$
Now $A$ has exactly the same number of three-term arithmetic progressions as $A-x$, so we can invoke
Lemma~{\densityincrement} again with $V$ in place of $G$ and $A-x$ in place of $A$, but note that $\alpha$
has been replaced by $(1+\eps/64)\alpha$, so this iteration can only happen
$\log_{1+\eps/64}(1/\alpha) = O\bigl(\log(1/\alpha)\bigr)$ times
before the second case of the lemma becomes impossible, since $\alpha\le 1$. Hence we deduce that
there is some subspace $V$ of codimension $O\bigl(\log(1/\alpha)^9\bigr)$ and some translate $A+x'$ of $A$
such that
$$\ex_{x\in V} \ex_{y\in V} \mu_{A'}(y)\mu_{A'}(x-y)\mu_{2\cdot A'}(x) \ge {1\over 2},$$
where $A' = (A+x)\cap V$ and
the relative densities are taken with respect to the subspace $V$. Expanding further, this implies that
$${|V|^3\over |A'|^3|V|^2} \sum_{x\in V}\sum_{y\in V}
\one_{A'}(y) \one_{A'}(x-y)\one_{2\cdot A'}(x) \ge {1\over 2};$$
that is,
$$\bigl| \{ (x,y)\in (2\cdot A) \times A : x-y\in A'\}\bigr| \ge {|A'|^3|V|^5\over 2|V|^3}
\ge {\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)}.$$
Renaming variables, this counts the number of pairs $(x,z) \in A'\times A'$ such that $x+z = 2y$ for some
$y\in A'$. Since this equation implies that $z-y = y-x$, the above expression
counts the number of three-term arithmetic
progressions in $A'$, including the $|A'|$ trivial instances of $x=y=z$. This proves the first part of the
theorem.

For the last part of the theorem statement,
suppose that $A$ does not contain any nontrivial three-term arithmetic progressions.
Then
$${\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)} \le \bigl| \{ (x,y)\in (2\cdot A) \times A : x-y\in A'\}\bigr|
\le |A'| \le \alpha q^n,$$
whence
$$q^n \le {2q^{O(\log(1/\alpha)^9)}\over \alpha^2},$$
and taking $q$th logs of both sides yields
$$n \le \log_q\biggl({2q^{O(\log(1/\alpha)^9)}\over \alpha^2}\biggr) = O\bigl(\log(1/\alpha)^9\bigr).$$
Letting $C$ be the constant implicit in the $O$-notation, we invert this to obtain the bound
$$\alpha\le e^{-n^{1/9}/C^{1/9}} = e^{-\Omega(n^{1/9})},$$
which is what we wanted.\slug
